Separating random and deterministic sources of computational noise in explore-exploit decisions

### 2025-04-22 Plos Comp Bio - revision v1/manuscript_DeterministicRandomNoise.tex
noise paper revision

\title{Separating random and deterministic sources of computational noise in explore-exploit decisions}
\author[1,\textcurrency]{Siyu Wang}
\author[1,2,3]{Robert C. Wilson}

\affil[1]{Department of Psychology, University of Arizona, Tucson AZ, USA}
\affil[2]{Neuroscience and Physiological Sciences Graduate Interdisciplinary Program, University of
	Arizona, Tucson AZ, USA}
\affil[3]{Cognitive Science Program, University of Arizona, Tucson AZ, USA}
\affil[ \textcurrency]{Current Address: Laboratory of Neuropsychology, National Institute of Mental Health, National Institutes of Health, Bethesda MD, USA}

@@

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, \added[id=siyu]{both theoretical work in machine learning and empirical human studies} suggest that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 14$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 86$\%$ of the variability is truly `random', although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}
	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

@@

	It has recently been shown that humans appear to use random exploration and can increase decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014}, \added[id=siyu]{as has also been suggested in computational models of animal behavior \citep{DOYA2002495, KHAMASSI2013441}}. In one of these tasks, known as the Horizon Task \citep{wilson2014}, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you will be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	One limitation of this previous research, however, is that it is difficult to tell whether what we have called `decision noise' actually reflect a noise process. From a modeling perspective, decision noise as defined in previous research essentially quantifies the extent to which behavior cannot be explained by a computational model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be random noise. For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take the ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness. \added[id=siyu]{To dissociate a missing deterministic component from a true random process, choice consistency between repeated decisions can be utilized to decompose behavioral variability into predictable deterministic components and unpredictable random components \citep{WYART2016, Wyart_2018, Findling19, Findling21}.}

@@

	\subsection*{Both behavioral variability and information seeking increase with horizon}
	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results using both a model-free and model-based approach \citep{wilson2014}. In both analyses, we focus on just the first free-choice trial in each game, where the only thing that differs between the horizon conditions is the number of choices that participants will make in the future. \added[id=siyu]{Subsequent choices in Horizon 6 games were not analyzed.}

@@

%\deleted[id=siyu]{For [1 3] condition, t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6; For [2 2] condition, t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6}
\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RanDetNoise_pinconsistent.jpg}
		\caption{Model-free analysis suggests that both deterministic and random noise contribute to the choice variability in random exploration. For both the [1 3] (A) and [2 2] (B) condition, people show greater choice inconsistency in horizon 6 than horizon 1. However, the extent to which their choices are inconsistent lies between what is predicted by purely deterministic and random noise, suggesting that both noise sources influence the decision.}
		\label{fig:mf2}
	\end{center}
\end{figure}

To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely deterministic and purely random noise cases.  For purely deterministic noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely random noise, the two games should be treated independently. \added[id=siyu]{Since repeated decisions mean that participants either choose the low-mean option twice, or choose the high-mean option twice, we could predict} the choice inconsistency, $p(\mbox{inconsistent})$, based on the probability of choosing the low mean option, $p(\mbox{low mean})$, as
\begin{equation*}
	\begin{split}
		p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
		&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
		\mbox{hence},\quad p(\mbox{inconsistent}) &=  
		1 - p(\mbox{consistent}) = 
		2 p(\mbox{low mean})(1-p(\mbox{low mean}))
	\end{split}
\end{equation*}

@@

For each pair of repeated games, the set of forced-choice trials are exactly the same, so the deterministic noise, $n_{det}$, should be the same while the random noise, $n_{ran}$ may be different. This is exactly how we distinguish deterministic noise from random noise. In symbolic terms, for repeated games $i$ and $j$,  $n_{det}^i=n_{det}^j$  and $n_{ran}^i \neq n_{ran}^j$. \added[id=siyu]{While an increase in either random or deterministic noises could lead to higher $p(\mbox{low mean})$, an increase in random noise predicts higher $p(\mbox{inconsistent})$ while an increase in deterministic noise predicts lower $p(\mbox{inconsistent})$. }

@@

\added[id=siyu]{We also fit a series of reduced and alternative models to the data. This includes reduced models that assume only deterministic or random noises. We also fit models in which the standard deviation of random and deterministic noises $\sigma_{det}$ and $\sigma_{ran}$ are estimated separately for [1 3] and [2 2] information conditions. Lastly, we fit a model with an alternative definition of $\Delta I$ that $\Delta I$ is defined to be difference between the variances of rewards shown on the forced-choice trials. Results on these model variants are presented in the Supplementary Materials.}

@@

\added[id=siyu]{Similar results are found in other variants of our model. Results for a model that estimates random and deterministic noises separately for [1 3] and [2 2] conditions, and a model with an alternative definition of information bonus $dI$, can be found in the Supplemental Materials (Supplemental Figure S12, S13). }

\added[id=siyu]{To ensure that the joint increase of random and deterministic noises is genuine and not an artifact from the fitting procedure, we computed the correlation between ground-truth values of random noise, and best-fitting values of deterministic noise (and vice versa), and they do not correlate (Supplementary Figure S14). Furthermore, we simulated data from a series of reduced models with known random and deterministic noise values in which either random or deterministic noise does not change with Horizon, and fit our model to the simulated data. Our model detects and only detects a change in random/deterministic noise with horizon, when the change is present in the model that simulates the data (Supplementary Figure S15, S16).}

@@

To perform this `posterior predictive check', we created a set of simulated data by taking the subject-level parameters from the hierarchical Bayesian fits and having the model play the same sequence of games as seen by the subjects. We then applied the same model-free analysis as described in the previous sections to this simulated data set and compared the model's behavior to that of participants. As shown in Figure \ref{fig:mb3}, the model can account for all qualitative patterns in the data --- the increase in p(high info), p(low mean), and p(inconsistent) with horizon, and that p(inconsistent) is in between pure random and pure deterministic noise. \added[id=siyu]{The quantitative agreement is almost perfect for p(high info) and for p(inconsistent), but the model slightly overestimate p(low mean) in [2 2] conditions. This has to do with the skewness of the subject-level posterior distribution (see Supplemental Materials). }
%The quantitative agreement is almost perfect for p(high info) and for p(inconsistent) in the [1 3] condition, but the model seems to systematically overestimate p(low mean) and p(inconsistent) in [2 2] conditions, although the discrepancy is relatively small (overestimating p(low mean) by 0.054 or 31.37\%, and p(inconsistent) by 0.049 or 27.83\% in [2 2] condition).

@@

Although the random noise interpretation is theoretically appealing, our approach, while an improvement on previous methods, is not without limitations. Most important is that our measure of `random' noise is only an upper bound on the true level of randomness and that, in principle, the random decision noise could be lower. Specifically, in our model, what we labeled random noise was really `non-stimulus-driven variability'. While this non-stimulus-driven variability could be driven by truly random stochastic processes, it could also be driven by deterministic processing that is unrelated to the stimuli in the task. For example, such deterministic noise could be driven by differences in where people look, or for how long they look, or by whether they were fidgeting or scratching their nose \citep{Musall2019}. \added[id=siyu]{Another limitation is that deterministic noise is defined based on the stimulus within a game, between-game deterministic strategies and stimuli from previous games (e.g., memory of previously seen game) were also treated as random noises in our model, although our model could be potentially extended by considering deterministic noise over a sequence of stimuli across games \citep{WYART2016, Wyart_2018}.} In addition to this conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also slightly underestimates deterministic noise (see Figure \ref{fig:paramrecover_main}, Supplementary Figure S5, S6). As a result, from both a conceptual and methodological perspective, it is possible that the remaining 86\% of the decision noise that is not stimulus-driven noise, could be deterministic.

@@

\added[id=siyu]{We show first that a change in noise is mathematically equivalent to a change in reward signal strength in our decision model (see also \cite{Cinotti19}).}. To show how a change in reward processing could affect random and deterministic noise, consider the simple decision model we introduced in Equation \ref{eq:newmodel}. In this model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, where
\begin{equation}
	\label{eq:noise1}
	\Delta Q = 
	\Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Now imagine a case where the reward signal is scaled by a factor $\beta$.  In this case, $\Delta Q$ becomes
\begin{equation}
	\label{eq:noise2}
	\Delta Q = 
	\beta \Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Because the choice only depends on the sign of $\Delta Q$, scaling $\Delta Q$ by a factor of $1/\beta$ will not change the behavior of the model.  Thus, if we divide both sides of the above equation by $\beta$ we get
\begin{equation}
	\label{eq:noise3}
	\Delta Q / \beta = 
	\Delta R + A \Delta I / \beta + b / \beta + n_{det} / \beta + n_{ran} / \beta
\end{equation}
which is equivalent to a scaling of both deterministic and random noise by the same factor $1/\beta$. Thus, one interpretation of our result that both deterministic and random noise change across horizons with the same ratio, is that this reflects a change in reward processing. That is, the reward signal is reduced in the longer horizon condition (smaller $\beta$ in horizon 6 than horizon 1).  %In the alternate interpretation, where deterministic and random noise are varied independently it is not clear why they should change by the same scale factor.

@@

% In particular, while we controlled many aspects of the stimulus across repeated games (e.g. the outcomes and the order of the forced trials), we could not perfectly control {\it all} stimuli the participant received, which would vary, for example, based on exactly what they were looking at or whether they were fidgeting or scratching their nose \citep{Musall2019}. Thus, our estimate of random noise is an upper bound as these `missing' sources of deterministic noise would be interpreted as random noise in our model. Conversely, our estimate of deterministic noise is a lower bound. \added[id=siyu]{Future work is needed to identify these additional sources of deterministic noise that are not controlled in our work, for example by tracking people's pose, body movements and eye movements during the experiment.}

@@

% \added[id=siyu]{The deterministic noise hypothesis is in line with works in which neural variability can be accounted for by fluctuations in sensory inputs. For example, }MT neurons were shown to have a reproducible temporal modulation in response to a fixed random motion stimuli \citep{Bair1996}. In other words, `irrelevant' features in the stimuli are represented in a reliable way in the brain that could drive downstream choices in a predictable way. Of particular interest in our study was the fact that deterministic noise increased with horizon. Such a horizon increase is a hallmark of an exploratory process and suggests that the modulation of deterministic processes may underlie random exploration.

% \added[id=siyu]{The random noise hypothesis, on the other hand, is consistent with findings of \cite{drugowitsch16}.} In particular these authors show that randomness in behavior could arise from imperfections in mental inference, which happen inside the brain, rather than in peripheral processes such as sensory processing and response selection. This suggests that a large proportion of variability in behavior may arise from computational errors in computing the correct strategy. \added[id=siyu]{Although suboptimal inference is different from simply adding random noise to neural circuitry\citep{Pouget12}, as long as the suboptimality in neural computation is not a deterministic function of the stimuli, it is a form of random noise in our definition. }In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute.	
%SIYU says the song bird literature supports that behavioral variability is reflected in neural variability. But it can't tell whether the neural variability arises from deterministic processes of external stimuli or random noise. I didn't read/remember the songbird paper, this is only based on the text below.
%"The idea that random exploration is driven by truly stochastic processes in the brain is in line with work in the bird song literature in which variability during song learning has been tied to neural variability arising from specific areas of the brain \citep{songbird1, songbird2}"

@@

%Taken at face value, the horizon-dependent increase in random noise is consistent with the idea that random exploration is driven by intrinsic variability in the brain. 
% \added[id=siyu]{
% % Regardless of whether the remaining 85\% is deterministic or random, the fact that the horizon change in the two noises are proportional to each other (Figure \ref{fig:ratio}) suggests a possible mechanism for random exploration. Specifically, a reduction in the strength with which reward drives the choice. In our decision model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, thus, the absolute value of different terms in the model do not matter, $\Delta Q$ is only affected by the relative magnitude of reward, information, bias and noise. Mathematically, our model is equivalent to
% % }
% \begin{equation*}
% 	\begin{split}
	% 		\Delta Q' &= \beta \Delta Q \\
	% 		&= \beta \Delta R + \beta A \Delta I + \beta b + \beta n_{det} + \beta n_{ran}\\
	% 		&= \beta \Delta R + A' \Delta I + b' + n_{det}' + n_{ran}'
	% 	\end{split}
% \end{equation*}
% \added[id=siyu]{In this equivalent form, $\beta$ determines the weight given to the reward in the decision. From a model fitting perspective, these two forms are equivalent. A decrease in $\beta$ here would be equivalent to an increase in variance of both deterministic and random noise. From a psychological perspective, however, they are quite different. An increase in both random and deterministic noise at a similar ratio relative to the reward, as observed in our experiment when horizon increases, suggests that either $\beta$ is reduced or $n_{det}'$ and $n_{ran}'$ are increased simultaneously at the same ratio. Neurally, reducing the $\beta$ corresponds to a reduction in reward sensitivity and reward coding, whereas simultaneously increasing $n_{det}'$ and $n_{ran}'$ suggests a common noise gating mechanism in which the noise filtering neural circuit is inhibited so that both random and deterministic noises are amplified. }
%SIYU says the Ebitz paper sounds like choice signal is reduced during exploration, and p(switch|reward outcome) increases behaviorally in exploration state compared to exploitation? I am not sure it supported reward reduction?
%Such reduced coding of reward would be consistent with \cite{ebitz17}.%In a recent report from \cite{ebitz17}, the behavioral variability of monkeys in an `explore' state was also tied to random rather than deterministic sources of noise.

@@

\subsection*{Data and code}
Behavioral data as well as MATLAB codes to recreate the main figures from this paper will be made available upon publication. %website at https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/CZT6EE.

>>>>

\title{Separating random and deterministic sources of computational noise in explore-exploit decisions}
\author[1]{Siyu Wang}%,\textcurrency
\author[1,2,3]{Robert C. Wilson}

\affil[1]{Department of Psychology, University of Arizona, Tucson AZ, USA}
\affil[2]{Neuroscience and Physiological Sciences Graduate Interdisciplinary Program, University of
	Arizona, Tucson AZ, USA}
\affil[3]{Cognitive Science Program, University of Arizona, Tucson AZ, USA}
%\affil[ \textcurrency]{Current Address: Laboratory of Neuropsychology, National Institute of Mental Health, National Institutes of Health, Bethesda MD, USA}

@@

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, \added{both theoretical work in machine learning and empirical human studies} suggest that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 14$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 86$\%$ of the variability is truly `random', although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}

	\newpage
	\section*{Author Summary}
	\added{
	Human decisions often seem random. Even simple decisions like what food to order at a restaurant can be difficult to predict ahead of time. This randomness in our decisions can be beneficial, effectively allowing us to explore new options. One outstanding question is where the randomness in our decisions comes from. Sometimes, our seemingly random decisions are driven by predictable external factors, like what the guest at the next table ordered could influence what we order. Other times, our decisions are not driven by external factors but are instead made by random thoughts within our brain. In this work, we developed a computational method that quantifies the extent to which the apparent randomness in our decisions can be explained by deterministic sources of variability in the external stimuli, or random variability unexplained by the stimuli. We found evidence that randomness in exploratory decisions can be explained by both random (up to 86\%) and deterministic (more than 14\%) sources of variability. Moreover, our results suggest that both sources of variability are adaptive, which enables humans to explore more when it is more beneficial to explore. The joint adaptation of random and deterministic noises also suggests a common noise-gating mechanism for exploration.  }

	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

@@

	It has recently been shown that humans appear to use random exploration and can increase decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014}, \added{as has also been suggested in computational models of animal behavior \citep{DOYA2002495, KHAMASSI2013441}}. In one of these tasks, known as the Horizon Task \citep{wilson2014}, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you will be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	One limitation of this previous research, however, is that it is difficult to tell whether what we have called `decision noise' actually reflect a noise process. From a modeling perspective, decision noise as defined in previous research essentially quantifies the extent to which behavior cannot be explained by a computational model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be random noise. For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take the ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness. \added{To dissociate a missing deterministic component from a true random process, choice consistency between repeated decisions can be utilized to decompose behavioral variability into predictable deterministic components and unpredictable random components \citep{WYART2016, Wyart_2018, Findling19, Findling21}.}

@@

	\subsection*{Both behavioral variability and information seeking increase with horizon}
	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results using both a model-free and model-based approach \citep{wilson2014}. In both analyses, we focus on just the first free-choice trial in each game, where the only thing that differs between the horizon conditions is the number of choices that participants will make in the future. \added{Subsequent choices in Horizon 6 games were not analyzed.}

@@

%\deleted{For [1 3] condition, t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6; For [2 2] condition, t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6}
\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RanDetNoise_pinconsistent.jpg}
		\caption{Model-free analysis suggests that both deterministic and random noise contribute to the choice variability in random exploration. For both the [1 3] (A) and [2 2] (B) condition, people show greater choice inconsistency in horizon 6 than horizon 1. However, the extent to which their choices are inconsistent lies between what is predicted by purely deterministic and random noise, suggesting that both noise sources influence the decision.}
		\label{fig:mf2}
	\end{center}
\end{figure}

To gain more quantitative insight into these results, we computed theoretical values for the choice inconsistency for the purely deterministic and purely random noise cases.  For purely deterministic noise this computation is simple because people should make the exact same decisions each time in repeated games, meaning that $p(\mbox{inconsistent}) = 0$ in this case. For purely random noise, the two games should be treated independently. \added{Since repeated decisions mean that participants either choose the low-mean option twice, or choose the high-mean option twice, we could predict} the choice inconsistency, $p(\mbox{inconsistent})$, based on the probability of choosing the low mean option, $p(\mbox{low mean})$, as
\begin{equation*}
	\begin{split}
		p(\mbox{consistent}) &= p(\mbox{low mean})^2 + p(\mbox{high mean})^2\\
		&= p(\mbox{low mean})^2 + (1-p(\mbox{low mean}))^2\\ 
		\mbox{hence},\quad p(\mbox{inconsistent}) &=  
		1 - p(\mbox{consistent}) = 
		2 p(\mbox{low mean})(1-p(\mbox{low mean}))
	\end{split}
\end{equation*}

@@

For each pair of repeated games, the set of forced-choice trials are exactly the same, so the deterministic noise, $n_{det}$, should be the same while the random noise, $n_{ran}$ may be different. This is exactly how we distinguish deterministic noise from random noise. In symbolic terms, for repeated games $i$ and $j$,  $n_{det}^i=n_{det}^j$  and $n_{ran}^i \neq n_{ran}^j$. \added{While an increase in either random or deterministic noises could lead to higher $p(\mbox{low mean})$, an increase in random noise predicts higher $p(\mbox{inconsistent})$ while an increase in deterministic noise predicts lower $p(\mbox{inconsistent})$. }

@@

\added{We also fit a series of reduced and alternative models to the data. This includes reduced models that assume only deterministic or random noises. We also fit models in which the standard deviation of random and deterministic noises $\sigma_{det}$ and $\sigma_{ran}$ are estimated separately for [1 3] and [2 2] information conditions. Lastly, we fit a model with an alternative definition of $\Delta I$ that $\Delta I$ is defined to be difference between the variances of rewards shown on the forced-choice trials. Results on these model variants are presented in the Supplementary Materials.}

@@

\added{Similar results are found in other variants of our model. Results for a model that estimates random and deterministic noises separately for [1 3] and [2 2] conditions, and a model with an alternative definition of information bonus $dI$, can be found in the Supplemental Materials (Supplemental Figure S12, S13). }

\added{To ensure that the joint increase of random and deterministic noises is genuine and not an artifact from the fitting procedure, we computed the correlation between ground-truth values of random noise, and best-fitting values of deterministic noise (and vice versa), and they do not correlate (Supplementary Figure S14). Furthermore, we simulated data from a series of reduced models with known random and deterministic noise values in which either random or deterministic noise does not change with Horizon, and fit our model to the simulated data. Our model detects and only detects a change in random/deterministic noise with horizon, when the change is present in the model that simulates the data (Supplementary Figure S15, S16).}

@@

To perform this `posterior predictive check', we created a set of simulated data by taking the subject-level parameters from the hierarchical Bayesian fits and having the model play the same sequence of games as seen by the subjects. We then applied the same model-free analysis as described in the previous sections to this simulated data set and compared the model's behavior to that of participants. As shown in Figure \ref{fig:mb3}, the model can account for all qualitative patterns in the data --- the increase in p(high info), p(low mean), and p(inconsistent) with horizon, and that p(inconsistent) is in between pure random and pure deterministic noise. \added{The quantitative agreement is almost perfect for p(high info) and for p(inconsistent), but the model slightly overestimate p(low mean) in [2 2] conditions. This has to do with the skewness of the subject-level posterior distribution (see Supplemental Materials). }
%The quantitative agreement is almost perfect for p(high info) and for p(inconsistent) in the [1 3] condition, but the model seems to systematically overestimate p(low mean) and p(inconsistent) in [2 2] conditions, although the discrepancy is relatively small (overestimating p(low mean) by 0.054 or 31.37\%, and p(inconsistent) by 0.049 or 27.83\% in [2 2] condition).

@@

Although the random noise interpretation is theoretically appealing, our approach, while an improvement on previous methods, is not without limitations. Most important is that our measure of `random' noise is only an upper bound on the true level of randomness and that, in principle, the random decision noise could be lower. Specifically, in our model, what we labeled random noise was really `non-stimulus-driven variability'. While this non-stimulus-driven variability could be driven by truly random stochastic processes, it could also be driven by deterministic processing that is unrelated to the stimuli in the task. For example, such deterministic noise could be driven by differences in where people look, or for how long they look, or by whether they were fidgeting or scratching their nose \citep{Musall2019}. \added{Another limitation is that deterministic noise is defined based on the stimulus within a game, between-game deterministic strategies and stimuli from previous games (e.g., memory of previously seen game) were also treated as random noises in our model, although our model could be potentially extended by considering deterministic noise over a sequence of stimuli across games \citep{WYART2016, Wyart_2018}.} In addition to this conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also slightly underestimates deterministic noise (see Figure \ref{fig:paramrecover_main}, Supplementary Figure S5, S6). As a result, from both a conceptual and methodological perspective, it is possible that the remaining 86\% of the decision noise that is not stimulus-driven noise, could be deterministic.

@@

\added{We show first that a change in noise is mathematically equivalent to a change in reward signal strength in our decision model (see also \cite{Cinotti19}).}. To show how a change in reward processing could affect random and deterministic noise, consider the simple decision model we introduced in Equation \ref{eq:newmodel}. In this model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, where
\begin{equation}
	\label{eq:noise1}
	\Delta Q = 
	\Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Now imagine a case where the reward signal is scaled by a factor $\beta$.  In this case, $\Delta Q$ becomes
\begin{equation}
	\label{eq:noise2}
	\Delta Q = 
	\beta \Delta R + A \Delta I + b + n_{det} + n_{ran}
\end{equation}
Because the choice only depends on the sign of $\Delta Q$, scaling $\Delta Q$ by a factor of $1/\beta$ will not change the behavior of the model.  Thus, if we divide both sides of the above equation by $\beta$ we get
\begin{equation}
	\label{eq:noise3}
	\Delta Q / \beta = 
	\Delta R + A \Delta I / \beta + b / \beta + n_{det} / \beta + n_{ran} / \beta
\end{equation}
which is equivalent to a scaling of both deterministic and random noise by the same factor $1/\beta$. Thus, one interpretation of our result that both deterministic and random noise change across horizons with the same ratio, is that this reflects a change in reward processing. That is, the reward signal is reduced in the longer horizon condition (smaller $\beta$ in horizon 6 than horizon 1).  %In the alternate interpretation, where deterministic and random noise are varied independently it is not clear why they should change by the same scale factor.

@@

% In particular, while we controlled many aspects of the stimulus across repeated games (e.g. the outcomes and the order of the forced trials), we could not perfectly control {\it all} stimuli the participant received, which would vary, for example, based on exactly what they were looking at or whether they were fidgeting or scratching their nose \citep{Musall2019}. Thus, our estimate of random noise is an upper bound as these `missing' sources of deterministic noise would be interpreted as random noise in our model. Conversely, our estimate of deterministic noise is a lower bound. \added{Future work is needed to identify these additional sources of deterministic noise that are not controlled in our work, for example by tracking people's pose, body movements and eye movements during the experiment.}

@@

% \added{The deterministic noise hypothesis is in line with works in which neural variability can be accounted for by fluctuations in sensory inputs. For example, }MT neurons were shown to have a reproducible temporal modulation in response to a fixed random motion stimuli \citep{Bair1996}. In other words, `irrelevant' features in the stimuli are represented in a reliable way in the brain that could drive downstream choices in a predictable way. Of particular interest in our study was the fact that deterministic noise increased with horizon. Such a horizon increase is a hallmark of an exploratory process and suggests that the modulation of deterministic processes may underlie random exploration.

% \added{The random noise hypothesis, on the other hand, is consistent with findings of \cite{drugowitsch16}.} In particular these authors show that randomness in behavior could arise from imperfections in mental inference, which happen inside the brain, rather than in peripheral processes such as sensory processing and response selection. This suggests that a large proportion of variability in behavior may arise from computational errors in computing the correct strategy. \added{Although suboptimal inference is different from simply adding random noise to neural circuitry\citep{Pouget12}, as long as the suboptimality in neural computation is not a deterministic function of the stimuli, it is a form of random noise in our definition. }In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute.	
%SIYU says the song bird literature supports that behavioral variability is reflected in neural variability. But it can't tell whether the neural variability arises from deterministic processes of external stimuli or random noise. I didn't read/remember the songbird paper, this is only based on the text below.
%"The idea that random exploration is driven by truly stochastic processes in the brain is in line with work in the bird song literature in which variability during song learning has been tied to neural variability arising from specific areas of the brain \citep{songbird1, songbird2}"

@@

%Taken at face value, the horizon-dependent increase in random noise is consistent with the idea that random exploration is driven by intrinsic variability in the brain. 
% \added{
% % Regardless of whether the remaining 85\% is deterministic or random, the fact that the horizon change in the two noises are proportional to each other (Figure \ref{fig:ratio}) suggests a possible mechanism for random exploration. Specifically, a reduction in the strength with which reward drives the choice. In our decision model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, thus, the absolute value of different terms in the model do not matter, $\Delta Q$ is only affected by the relative magnitude of reward, information, bias and noise. Mathematically, our model is equivalent to
% % }
% \begin{equation*}
% 	\begin{split}
	% 		\Delta Q' &= \beta \Delta Q \\
	% 		&= \beta \Delta R + \beta A \Delta I + \beta b + \beta n_{det} + \beta n_{ran}\\
	% 		&= \beta \Delta R + A' \Delta I + b' + n_{det}' + n_{ran}'
	% 	\end{split}
% \end{equation*}
% \added{In this equivalent form, $\beta$ determines the weight given to the reward in the decision. From a model fitting perspective, these two forms are equivalent. A decrease in $\beta$ here would be equivalent to an increase in variance of both deterministic and random noise. From a psychological perspective, however, they are quite different. An increase in both random and deterministic noise at a similar ratio relative to the reward, as observed in our experiment when horizon increases, suggests that either $\beta$ is reduced or $n_{det}'$ and $n_{ran}'$ are increased simultaneously at the same ratio. Neurally, reducing the $\beta$ corresponds to a reduction in reward sensitivity and reward coding, whereas simultaneously increasing $n_{det}'$ and $n_{ran}'$ suggests a common noise gating mechanism in which the noise filtering neural circuit is inhibited so that both random and deterministic noises are amplified. }
%SIYU says the Ebitz paper sounds like choice signal is reduced during exploration, and p(switch|reward outcome) increases behaviorally in exploration state compared to exploitation? I am not sure it supported reward reduction?
%Such reduced coding of reward would be consistent with \cite{ebitz17}.%In a recent report from \cite{ebitz17}, the behavioral variability of monkeys in an `explore' state was also tied to random rather than deterministic sources of noise.

@@

\subsection*{Data and code}
Behavioral data as well as MATLAB codes to recreate the main figures from this paper will be made available upon publication at \url{https://github.com/wangxsiyu/RW_RandomDeterministicNoise.git}

\section*{Supporting information}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Replication of previous findings with data from all participants (i.e. no exclusions).} (A) model-free measure of behavioral variability, $p(\mbox{low mean})$, increases with horizon. (B) model-free measure of information seeking, $p(\mbox{high info})$, increases with horizon. (C) model-based measure of behavioral variability, decision noise $\sigma$, increases with horizon. (D) model-based measure of information seeking, information bonus $A$, increases with horizon.

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf Model-free analysis with data from all participants (i.e. no exclusions) suggests that both deterministic and random noise contribute to the choice variability in random exploration. } For both the [1 3] (A) and [2 2] (B) condition, people show greater choice inconsistency in horizon 6 than horizon 1. However, the extent to which their choices are inconsistent lies between what is predicted by purely deterministic and random noise, suggesting that both noise sources influence the decision..
